Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2018-11-02 scagnostics.Rmd
plots of data sets

---
title: An investigation into using scagnostics to find interesting projections of high-dimensional data

@@

keywords:
- statistical graphics
- data visualisation
- exploratory data analysis
- data science 
- guided tour

@@

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the scagnostics on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $p$ is used for the number of points, $n$ is the number of dimensions, and $d=2$ is the projection dimension.

@@

Two samples sizes are also used: $n=(100, 1000)$.

@@

To understand the potential for the various indices defined we use the following overview procedures on the simulated data sets:

- index values for pair of variables, explicitly comparing the uninformative and special view of each dataset, as well as discussing the dependence on the sample size, see Section \ref{sec:dataOv}
- tour between pairs of variables, considering a planned tour that moves from the $x_1 - x_2$ view either to the $x_3 - x_4$ view (no special structure encountered over the full path) or to the $x_5 - x_6$ (moving from uninformed to special view), see Section \ref{sec:smooth}
- optimisation path, starting from $x_1 - x_2$, how can we use the available index functions and optimisation algorithms to identify the special view, see Section \ref{sec:guided}
- rotating a dataset with functional dependence to evaluate the rotation dependence of the index functions, see Section \ref{sec:rot}
- benchmarking the computing time in Section \ref{sec:timer}

\textcolor{red}{\cite{posse95a} uses simulations with "strongly hidden structure", i.e. truncated Gaussian distributions with approximately identity covariance matrix, such that they don't appear in any of the principal component planes. In particular seems that spiral is hardest to identify based on results from Posse, maybe I can add this to my sphere dataset? or instead of it?}

### Dataset overview
\label{sec:dataOv}

To test the behaviour of the index functions we use three types of simulated data sets based on different underlying distributions, and with different built-in features. They are defined in $n=6$ dimensions with $p$ randomly selected points $x_i, i = 1,..,n$ according to:

- spiral:  XXX $p$ points on an n-dimensional sphere randomly selected using the geozoo package [@geozoo], as an example of dataset without special views, labelled "sphere"
- pipe: $p$ points where $n-1$ points are independently drawn from a uniform distribution between $[-1,1]$, and enforcing $x_n^2 + x_{n-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- sine:  $p$ points where $n-1$ points are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_n = \sin(x_{n-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"

For each distribution we consider a small sample of $p=100$ and a large sample with $p=1000$.

@@

```{r datasets}
set.seed(1984)
spiral100 <- spiralData(6, 100) %>% scale() %>% as_tibble()
spiral1000 <- spiralData(6, 1000) %>% scale() %>% as_tibble()
#sphere100 <- sphereData(6, 100) %>% scale() %>% as_tibble()
#sphere1000 <- sphereData(6, 1000) %>% scale() %>% as_tibble()

To get an overview of the dataset we show views of the first two directions showing the independent distribution and the last two directions showing the special view. The enforced relationships will also affect the 1-d densities as can be seen from \ref{fig:dataPlotsDensity}, which means that views including sizable fractions of $x_6$ (and $x_5$ for the Pipe and Spiral distribution) will be distinct from uninformed views of distributions 2 and 3.

```{r dataPlotsV1V2, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the first two parameters $x_1$ and $x_2$, i.e. showing a typical uninformed view for each dataset."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsV5V6, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the last two parameters $x_5$ and $x_6$, i.e. showing the special view built into the datasets."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V5, V6)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

>>>>

---
title: Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

@@

keywords:
- scagnostics
- statistical graphics
- data visualisation
- exploratory data analysis
- data science 
- guided tour

@@

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $p$ is used for the number of points, $n$ is the number of dimensions, and $d=2$ is the projection dimension.

@@

The structured projection is in fifth and sixth variables. Two samples sizes are also used: $n=(100, 1000)$. All variables are standardized to mean 0 and standard deviation 1.

```{r datasets}
# sample(1000:9999, 1)
set.seed(3705)
spiral100 <- spiralData(6, 100) %>% scale() %>% as_tibble()
spiral1000 <- spiralData(6, 1000) %>% scale() %>% as_tibble()
#sphere100 <- sphereData(6, 100) %>% scale() %>% as_tibble()
#sphere1000 <- sphereData(6, 1000) %>% scale() %>% as_tibble()

Figure \ref{data} shows samples from each of the families, both nuisance and structured projections.

<!--
To get an overview of the dataset we show views of the first two directions showing the independent distribution and the last two directions showing the special view. The enforced relationships will also affect the 1-d densities as can be seen from \ref{fig:dataPlotsDensity}, which means that views including sizable fractions of $x_6$ (and $x_5$ for the Pipe and Spiral distribution) will be distinct from uninformed views of distributions 2 and 3.
-->

```{r data, fig.width=6, fig.height=7, fig.cap="Scatterplots of pairs of variables from samples of each family, showing the nuisance variables and structured variables."}
dsn <- spiral1000 %>% select(V1, V2) %>%
  rename(X1=V1, X2=V2) %>%
  mutate(family="spiral", type="nuisance")
dss <- spiral1000 %>% select(V5, V6) %>%
  rename(X1=V5, X2=V6) %>%
  mutate(family="spiral", type="structured")
dpn <- pipe1000 %>% select(V1, V2) %>%
  rename(X1=V1, X2=V2) %>%
  mutate(family="pipe", type="nuisance")
dps <- pipe1000 %>% select(V5, V6) %>%
  rename(X1=V5, X2=V6) %>%
  mutate(family="pipe", type="structured")
din <- sin1000 %>% select(V1, V2) %>%
  rename(X1=V1, X2=V2) %>%
  mutate(family="sine", type="nuisance")
dis <- sin1000 %>% select(V5, V6) %>%
  rename(X1=V5, X2=V6) %>%
  mutate(family="sine", type="structured")
d <- bind_rows(dsn, dss, dpn, dps, din, dis)
ggplot(d, aes(x=X1, y=X2)) + geom_point(alpha=0.3) + 
  facet_grid(type~family) + theme(aspect.ratio=1) +
  xlab("") + ylab("")
```

@@

The family of data sets can be used to assess the PPIs on the desirable properties: smoothness, squintability, flexibility, rotation invariance, and speed. The procedures are as follows:

- Compare the index values for structured vs nuisance variables, by simulating many examples from the model families, and recording the 90% confidence intervals. This checks if the PPI can detect the structure, that it has distinct differences between the sets of values. 
- Use the tour to show the index values along an interpolation between pairs of nuisance variables, $x_1 - x_2$ view either to the $x_3 - x_4$. The result is ideally a smooth change in low values. This checks the smoothness property. 
- Use the tour to show the index values along an interpolation between a pair of nuisance variables $x_1 - x_2$ and the structured pair of variables $x_5 - x_6$. This examines the squintability, and smoothness. If the function is smooth and slowly increases towards the structured projection, then the structure is visible from a distance. 
- A guided tour, showing the optimisation path, starting from $x_1 - x_2$, for one index, shows how effectively the maximum is attained. The values for other PPIs is computed for this path also, which allows an assessment of similarity between PPIs. 
- Rotation invariance is checked by computing PPIs on rotations of the structured projection. 
- Benchmarking is used to assess the computational speed.

<!--
\textcolor{red}{\cite{posse95a} uses simulations with "strongly hidden structure", i.e. truncated Gaussian distributions with approximately identity covariance matrix, such that they don't appear in any of the principal component planes. In particular seems that spiral is hardest to identify based on results from Posse, maybe I can add this to my sphere dataset? or instead of it?}
-->

@@

```{r dataPlotsV1V2, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the first two parameters $x_1$ and $x_2$, i.e. showing a typical uninformed view for each dataset.", eval=FALSE}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsV5V6, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the last two parameters $x_5$ and $x_6$, i.e. showing the special view built into the datasets.", eval=FALSE}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V5, V6)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```
